Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-10 paper.tex
A few minor fixes

In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools such as
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and others. Such workflow engines usually
describe applications in a high-level language with specific data and
control flow constructs, parallelization operators, visual edition
tools, links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking, \closednote{Naj}{should we not mention PMP which is used for
  CIVET?}{Tristan}{Done, as discussed.}  redundancy detection across
analyses to avoid re-computation, and so on. A wide array of workflows
have been implemented using these pipelines and are now shared across
neuroimaging groups world-wide, which represents a tremendous
opportunity for science gateways to leverage. Domain-specific engines
nicely complement e-Science systems that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

@@

Scalability is also a crucial concern for such multi-user,
high-throughput systems. To this end, science gateways may balance the
load among different instances of the same engine, and use advanced
task scheduling policies on the infrastructure to improve performance,
fault-tolerance and fairness among users.

@@

Specific features may also be available, for instance
data vizualisation and quality control, workflow edition, debugging
instruments, social tools among users, etc.

@@

Architectures to integrate workflow engines in science gateways are
diagrammed in Figure~\ref{fig:architectures} using the graphical
notations shown in Figure~\ref{fig:notations}. This section describes the different types of
interactions involved in the architectures. Architectures are then
detailed and illustrated using examples of real
systems. Table~\ref{table:system-classification} summarizes the
classification of a few systems by architecture.

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored\closednote{sorina}{VIP and Moteur also share the
  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
  such abstract architectures may not completely apply in real
  life. See comment about ``blended'' architectures in the
  discussion. In case of VIP, the sharing of the DB is more an
  implementation detail rather than a strong architectural pattern: we
  could avoid sharing the DB if we wanted.}.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE~\cite{maddineni2012distributed}), in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
a particular task that can submit sub-tasks to the science gateway
through interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks, but their execution is delegated
to the science gateway that executes them on the infrastructure
through interaction \texttt{b}. Although the science gateway has no
global vision of the workflow, it can keep track of the sub-tasks
submitted by a given task, for instance to cancel them when
the task is canceled. The science gateway may also implement
mechanisms to facilitate the handling of task dependencies, for
instance basic dependency lists as available in most batch managers
(see for instance attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-FSL.pdf_tex}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}.]{\def\svgwidth{\columnwidth}
  \input{figures/CBRAIN-PSOM.pdf_tex
}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler. FSL sub-tasks will run locally
  instead of being submitted again to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

\subsection{Pool model}
\label{sec:pool}
See Figure~\subref*{archi:agent}. Workflows are submitted by the science
gateway to a central pool through interaction \texttt{d}. Agents connect to
the pool asynchronously to retrieve and execute workflows through
interaction \texttt{f}. Agents may be started according to various
policies, for instance to ensure load balancing. Agents may wrap
different types of workflow engines. Workflow engine controls tasks
and data on the infrastructure through interactions \texttt{b} and
\texttt{c$_2$}, science gateway transfers user data through
interaction \texttt{c$_1$}, and administrator installs workflows
through interaction \texttt{a$_1$} and \texttt{a$_2$}.

@@

See Figure~\subref*{archi:nested}. In nested workflows
(Figure~\subref*{archi:nested}-Left), workflows include other
workflows that are executed by a \emph{child} engine. However, parent
and child engines might use different languages and might run on
different infrastructures. A parent workflow is
also called meta-workflow. The science gateway communicates with the
parent engine through abstract interaction \texttt{*$_1$}. The science
gateway also communicates with the infrastructure to transfer user
data through abstract interactions \texttt{*$_2$} and
\texttt{*$_3$}. Both workflow engines communicate with the
infrastructure through abstract interactions \texttt{*$_4$} and
\texttt{*$_5$}. The parent engine communicates with the child engine
through abstract interaction \texttt{*$_6$}. Administrator installs
workflows through interactions \texttt{a$_1$} and \texttt{a$_2$}.

@@

\emph{Integration complexity} is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), which is
why we keep it in the metric. The two metrics are summed to obtain a
global score that measures the
total number of software pieces to develop, including interactions and
components. It may be used to measure the development effort required
to build the architecture.

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity since complex architectures tend to be more prone to
failure. More precisely, robustness is determined as the number of
software components and interactions that are specific to workflow
execution:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are the ones represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. 
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. \closednote{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics}{Tristan}{Metrics are compared separately too, see Table 2 and discussion.} Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

@@

Tight integration and task encapsulation are the simplest
architectures to integrate, followed by service integration, workflow import (with
service invocation) and pool. Nested workflows (with service
invocation) requires more integration than the other
architectures. Robustness roughly leads to the same ordering of
architectures, with tight integration and task encapsulation in the top
group, service integration and workflow import (with service
invocation) close behind, pool in a third group, and nested workflows
(with service invocation) at the end. This ordering is consistent
across metrics; it reflects the global complexity of the
architectures.

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which is
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Task encapsulation and workflow import (with service invocation)
perform a bit better when integrating new engines (\texttt{E$_1$}),
which is interesting given the current profusion of engines with
different characteristics.  However, task encapsulation, workflow import
(with service invocation) and tight integration perform worse than the
others when adding engine versions (\texttt{E$_2$}) due to the need to
update infrastructure (task encapsulation), science gateway (tight
integration), or workflow converter (workflow import). Workflow import
performs worse than the others when integrating new workflows
(\texttt{E$_3$}) because of the language conversion step. All
architectures except nested workflows perform the same when
integrating new infrastructures (\texttt{E$_4$}).

@@

A few limitations remain that should be considered when using the
results of our evaluation. First of all, the evaluation is done at a
quite high-level of abstraction that may be a bit artificial when
applied to real systems. Abstracting architectures from their
implementation in specific systems is useful to compare them, but we
also realize that it comes at the cost of realism. In particular, the
presented architectures are theoretical patterns that may be blended
in actual systems. The distinction between tight integration and
service invocation, for instance, may not always be that clear in
practice. Service invocation may also be combined with task
encapsulation in some cases. However, the criteria discussed in this
work would still apply to broadly compare and categorize such cases of
hydrid architectures.

@@

% Impact of the work
To the best of our knowledge, our work is the first of its kind. So
far, the literature on science gateways and workflow engines has
focused on the description of particular systems, or on the
presentation of a particular architecture. For instance, the work
in~\cite{Gannon2005} discusses service-oriented architectures for
science gateways with workflow engines.  Instead, our analysis
abstracts and evaluates architectural patterns independently from any
particular system. It provides general insight to integrate science
gateways and workflow engines, which we hope will be useful for
software architects.

% Literature on SG architecture Very few works go beyond the
% description of architectures of particular systems. The work in
% http://link.springer.com/chapter/10.1007/11596141_3 discusses
% service-oriented architectures for science gateways, with a workflow
% engine. The work in
% http://docs.lib.purdue.edu/cgi/viewcontent.cgi?article=1414&context=nanopub
% highlights 4 TeraGrid science gateways (GridChem, LEAD, nanoHUB,
% CABIG). Architectures are described but there is no real focus on
% workflow engines. GridChem doesn't include any workflow engine,
% which btw make it relevant for integration via sub-tasking. LEAD has
% a workflow engine.

>>>>

In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools among which
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing}
and WS-PGRADE~\cite{Kacsuk2012}. Such workflow engines usually
describe applications in a high-level language with specific data and
control flow constructs, parallelization operators, visual edition
tools, links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), 
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), RPPL~\cite{1174106}, SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking, \closednote{Naj}{should we not mention PMP which is used for
  CIVET?}{Tristan}{Done, as discussed.}  redundancy detection across
analyses to avoid re-computation, and so on. A wide array of workflows
have been implemented using these pipeline systems and are now shared across
neuroimaging groups world-wide, which represents a tremendous
opportunity for science gateways to leverage. Domain-specific engines
nicely complement e-Science systems that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

@@

Scalability is also a crucial concern for such multi-user,
high-throughput systems. For this purpose, science gateways may
balance the load among different instances of the same engine, start
new engines elastically, and use advanced task scheduling policies on
the infrastructure to improve performance, fault-tolerance and
fairness among users.

@@

Other specific features may also be available, for instance
data vizualisation and quality control, workflow edition, debugging
instruments, social tools among users, etc.

@@

Architectures to integrate workflow engines in science gateways are
diagrammed in Figure~\ref{fig:architectures} using the graphical
notations shown in
Figure~\ref{fig:notations}. Table~\ref{table:system-classification}
summarizes the classification of a few systems by architecture.

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a model
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where applications, users and other
resources are stored\closednote{sorina}{VIP and Moteur also share the
  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
  such abstract architectures may not completely apply in real
  life. See comment about ``blended'' architectures in the
  discussion. In case of VIP, the sharing of the DB is more an
  implementation detail rather than a strong architectural pattern: we
  could avoid sharing the DB if we wanted.}.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE)~\cite{maddineni2012distributed}, in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
a particular task that can submit sub-tasks to the science gateway
through interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks, but their execution is delegated
to the science gateway that executes them on the infrastructure
through interaction \texttt{b}. Although the science gateway has no
global vision of the workflow, it can keep track of the sub-tasks
submitted by a given task, for instance to cancel them when
the task is canceled. The science gateway may also implement
mechanisms to facilitate the handling of task dependencies, for
instance basic dependency lists as available in most cluster managers
(see for instance attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-FSL.pdf_tex}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}.]{\def\svgwidth{\columnwidth}
  \input{figures/CBRAIN-PSOM.pdf_tex
}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler (FSL sub-tasks or PSOM workers). FSL sub-tasks will run locally
  instead of being submitted again to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

\subsection{Pool model}
\label{sec:pool}
See Figure~\subref*{archi:agent}. Workflows are submitted by the
science gateway to a central pool through interaction
\texttt{d}. Agents connect to the pool asynchronously to retrieve and
execute workflows through interaction \texttt{f}. Agents may be
started according to various policies, for instance to ensure load
balancing. Workflow engine controls tasks and data on the
infrastructure through interactions \texttt{b} and \texttt{c$_2$},
science gateway transfers user data through interaction
\texttt{c$_1$}, and administrator installs workflows through
interaction \texttt{a$_1$} and \texttt{a$_2$}.

@@

See Figure~\subref*{archi:nested}. In nested workflows
(Figure~\subref*{archi:nested}-Left), workflows include other
workflows that are executed by a \emph{child} engine. Parent
and child engines might use different languages and might run on
different infrastructures. A parent workflow is
also called meta-workflow. The science gateway communicates with the
parent engine through abstract interaction \texttt{*$_1$}. The science
gateway also communicates with the infrastructure to transfer user
data through abstract interactions \texttt{*$_2$} and
\texttt{*$_3$}. Both workflow engines communicate with the
infrastructure through abstract interactions \texttt{*$_4$} and
\texttt{*$_5$}. The parent engine communicates with the child engine
through abstract interaction \texttt{*$_6$}. Administrator installs
workflows through interactions \texttt{a$_1$} and \texttt{a$_2$}.

@@

\emph{Integration complexity} is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), which is
why we keep it in the metric. The two metrics are summed to obtain a
global score that measures the
total number of software pieces to develop, including interactions and
components.

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity since complex architectures tend to be more prone to
failure. More precisely, robustness is determined as the number of
software components and interactions that are specific to workflow
execution:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are the ones represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down to 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. 
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. \closednote{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics}{Tristan}{Metrics are compared separately too, see Table 2 and discussion.} Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

@@

Tight integration and task encapsulation are the simplest
architectures to integrate, followed by service integration, workflow import (with
service invocation) and pool. Nested workflows (with service
invocation) require more integration than the other
architectures. Robustness roughly leads to the same ordering of
architectures, with tight integration and task encapsulation in the top
group, service integration and workflow import (with service
invocation) close behind, pool in a third group, and nested workflows
(with service invocation) at the end. This ordering is consistent
across metrics; it reflects the global complexity of the
architectures.

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which is
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Task encapsulation and workflow import (with service invocation)
perform a bit better when integrating new engines (\texttt{E$_1$}),
which is interesting given the current profusion of engines with
different characteristics.  However, task encapsulation, workflow import
(with service invocation) and tight integration perform worse than the
others when adding engine versions (\texttt{E$_2$}) due to the need to
update infrastructure (task encapsulation), science gateway (tight
integration), or workflow converter (workflow import). Workflow import
performs worse than the others when integrating new workflows
(\texttt{E$_3$}) because of the language conversion step. All
architectures except nested workflows are equivalent when
integrating new infrastructures (\texttt{E$_4$}).

@@

A few limitations should be considered when using the
results of our evaluation. First of all, the evaluation is done at a
quite high-level of abstraction that may be a bit artificial when
applied to real systems. Abstracting architectures from their
implementation in specific systems is useful to compare them, but we
also realize that it comes at the cost of realism. In particular, the
presented architectures are theoretical patterns that may be blended
in actual systems. The distinction between tight integration and
service invocation, for instance, may not always be that clear in
practice. Service invocation may also be combined with task
encapsulation in some cases. However, the criteria discussed in this
work would still apply to broadly compare and categorize such cases of
hydrid architectures.

@@

% Impact of the work
To the best of our knowledge, our work is the first of its kind. So
far, the literature on science gateways and workflow engines has
focused on the description of particular systems, or on the
presentation of a particular architecture.  Instead, our analysis
abstracts and evaluates architectural patterns independently from any
particular system. It provides general insight to integrate science
gateways and workflow engines, which we hope will be useful for
software architects.

% Literature on SG architecture Very few works go beyond the
% description of architectures of particular systems. For instance, the work
%in~\cite{Gannon2005} discusses service-oriented architectures for
% science gateways with workflow engines.  The work in
% http://link.springer.com/chapter/10.1007/11596141_3 discusses
% service-oriented architectures for science gateways, with a workflow
% engine. The work in
% http://docs.lib.purdue.edu/cgi/viewcontent.cgi?article=1414&context=nanopub
% highlights 4 TeraGrid science gateways (GridChem, LEAD, nanoHUB,
% CABIG). Architectures are described but there is no real focus on
% workflow engines. GridChem doesn't include any workflow engine,
% which btw make it relevant for integration via sub-tasking. LEAD has
% a workflow engine.
